Large scale risks from pesticides in small streams

### 2016-10-02 article.tex
more revisions

%% -------------------------------------------------------------------------
\section{Methods}
\subsection{Data compilation}
We compiled pesticide monitoring data from sampling sites with catchment sizes $\mathrm{< 150km^2}$ for the years 2005 to 2015 from all 13 non-city federal states of Germany (see Supplemental Table~S1 for the abbreviations of federal state names). 
% 150 km2 needs justification - can perhaps be linked to introduction, see comment above
We homogenized and unified all data provided by the federal states into a database and implemented a robust data cleaning workflow (see Supplemental Figure~S1 for details) \citep{poisot_best_2015}.

We identified chemical samples taken during heavy rainfall events by spatio-temporal intersection of sampling events with gridded daily precipitation data available from the German Weather service (DWD).
This data spatially interpolates daily precipitation values from local weather stations \citep{rauthe_central_2013}. 
We performed the intersection for the actual sampling date and the day before, to identify rain events during and up to 48 hours before sampling.
%details required - what is a heavy rainfall event, resolution of data. One question with such data is from which precipitation value increases in concentrations are detectable. Also interesting

\subsection{Characterization of catchments}
We compiled a total of 3,049 sampling sites with pesticide measurements.
We delineated catchments upstream for each of the sampling sites using a digital elevation model (DEM) \citep{eea_digital_2013} and the multiple flow direction algorithm \citep{holmgren_multiple_1994} as implemented in GRASS GIS 7 \citep{neteler_grass_2012}.
Catchment delineation was manually checked for accuracy by comparison with state stream networks.
Given that the delineated catchments were accurate only for 30\% of the sites,
%a reviewer might ask how accuracy was defined/cehcked
we compiled complementary catchment size data from authorities that provided catchments for an additional (60\% of sites). Ten \% of sites lacked catchment size data and were omitted from the operations outlined below.

For each catchment we calculated the relative cover (in \%) with agricultural area based on Official Topographical Cartographic Information System (ATKIS) of the land survey authorities \citep{adv_atkis_2016}.
%it may be worth mentioning, what agricultural land cover is
Additionally, we used agricultural cover data provided by authorities (19\% of sites). 
For 78\% of the sites both, the proportion of agricultural land use and catchment size were available.
% see do_overview.R for numbers
% this section is puzzling - if you have 90% catchments, why is it only 78% in the end? It is also unclear what is the ATKIS coverage and why additional state-wide??? cover data was used...

A clear definition of SWB in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
The WFD defines SWB with a catchment size between 10 and 100 km\textsuperscript{2}, without further categorisation of streams \textless 10km\textsuperscript{2}. 
\citet{lorenz_specifics_2016} defines SWB with catchment size \textless 10km\textsuperscript{2} as SWB.
Because of data scarcity of streams \textless 10~km\textsuperscript{2} (Figure~\ref{fig:fig3}) we define in this study all streams below 25~km\textsuperscript{2} as SWB. This catchment size corresponds to a stream width of approximately 2~meters (see Supplemental Figure~S2).
% erstens muss diese definitionsfragen weiter nach oben, da schon über die SWB gesprochen wird. Zweitens braucht hier eigentlich keine Definition zu erfolgen - man kann eher untersuchen, ob es eine zusammmenhang zwischen belastung und catchment size gibt. War das nicht früher mal drinne. Dann könnte man auch in der Einleitung diskutieren, dass das ein weiteres Ziel der Arbeit war. Drittens wundert man sich, warum wir Gewässer < 150 analysieren, wenn wir als Definition hier 25 anlegen.

\subsection{Characterization of pesticide pollution}
We characterised pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorization as part of the ecological risk assessment.
No unacceptable ecological effect are expected if the environmental concentration remains below this concentration. 
%state here sth. like: that Stehle and Schulz showed that RAC exceedances reflect a decrease in biodiversity and from this perspective are ecologically relevant indicators. Otherwise regulatory values may not be "real" trhesholds because politically influenced and this potential criticism should be taken ahead
The German Federal Environmental Agency (UBA) provided RACs for the 105 compounds with highest detection rates (Supplemental Table~S2). 
We expressed RACs as Risk Quotient (RQ):

@@

where $No_i$ is the observed number of exceedances at site $i$. 
% die normalen Leser werden SChwierigkeit haben, das zu verstehen, z.B. wo in der FOrmel überhaupt RQ > 1 auftaucht
We modeled $No_i$ as resulting from a negative binomial distribution ($NB$) with .....
The proportion of agriculture within the catchment ($agri_i$) and the catchment size of the site ($size_i$) were used as predictors of .... 
$f_1$ and $f_2$ are smoothing functions using penalized cubic regression splines \citep{wood_generalized_2006}.
The degree of smoothness was estimated using restricted maximum likelihood (REML) during model fitting process \citep{wood_fast_2011}.
The number of samples per site ($n_i$) was used as an offset to account for differences in sampling efforts (number of samplings and of analysed compounds) at a site and is equivalent to modeling the rate of exceedances. 
%bleibt unklar, wie beides sampling and compound spectrum in n eingeht
% beta 0 nicht erklärt - warum eigentlich 0 und nihct i, müssen alle den gleichen parameter haben - und macht es überhaupt sinn einen achsen-schnittpunkt ungleich 0 zu haben?
We used point-wise 95\% Confidence Intervals (CI) of the first derivative of the fitted smooth to check if there are regions of statistically significant changes.
GAMs were fitted using the mgcv package \citep{wood_fast_2011}.

While agricultural land use and catchment size vary only between sites, this is not the case for precipitation which changes also with time.
Therefore, we modeled the effects of precipitation in a separate model.
% begründung bleibt unklar - warum separates modell, denn im modell oben taucht weder raum noch hier zeit auf... Entscheidend ist doch auch, dass oben die Anzahl der RAC exdeendances und hier RQ Response variables sind
RQ and concentrations show a skewed distribution with an excess of zeros (no pesticides detected and quantified). 
Therefore, we modeled these as two processes 
% das gleiche hätte man oben auch shcon erklären können?
(one generating values below the limit of quantification (LOQ) and one generating values above LOQ) using a Zero-Adjusted Gamma (ZAGA) distribution \cite{rigby_generalized_2005,stasinopoulos_gamlss.dist:_2016}:

@@

$\nu_i$ denotes the probability of an observation i being above LOQ and $f_{Gamma}$ denotes the gamma function and is used for values equal to or greater LOQ, with $\mu$ being the mean and $\sigma$ the standard deviation.
%weiß nicht, ob der normale leser kapiert, dass das mean und sd von RQ sind...
We used the $log(x+0.05)$ transformed precipitation at sampling date ($log~prec_0$) and the day before ($log~prec_{-1}$), as well as quarters of the year ($Q1-Q4$) as linear predictors for $\mu$ and $\nu$. 
We used appropriate link functions for $\mu$ and $\nu$ and assumed $\sigma$ to be constant. 
Equation~\ref{eqn:eqn4} summarises the deterministic part of the model.

@@

We fitted this model separately to each compound with a RAC, measured in least 1000 samples and more than 5\% of values above LOQ (n = 24 compounds, see Supplemental Table~S3 for a list of compounds). 
To summarise the coefficients across the modeled compounds we used a random effect meta-analysis \citep{harrison_getting_2011}.

@@

The compiled dataset comprised only few standing waters (58 sites) and the majority of samples (91\%) where taken via grab sampling.  % see clean.R for numbers
9\% of samples from 33 sites were taken as composite samples of different durations.
Therefore, we restricted the analyses to grab samples from streams. 
The analyzed dataset comprised 2,918,604 pesticide measurements of 42,236 samples in 3,049 sampling sites.  %see do_overview.R for numbers.
We found large differences in the number of sampling sites between federal states and their spatial distribution (Figure~\ref{fig:fig1} and Supplemental Table~S1).

@@

The distribution of sampling sites across catchment area and agricultural area in the catchment revealed a sharp decline in the distribution of catchment-sizes below $10~km^2$, with most sampling sites with catchments between 10 and 25 $km^2$ (Figure~\ref{fig:fig3}).
The proportion of agriculture in the catchments decreased with increasing catchment size.

@@

\subsection{Influence of agricultural land use and catchment size}
Modeling the number of RAC exceedances as function of agriculture within catchment and catchment size revealed that there is a strong and statistically significant increase up to 25\% agriculture.
Above this threshold the exceedances level off followed by a increase above 75\% (Figure~\ref{fig:fig4}, left).
We could no detect any effect of catchment size on the number of RAC exceedances (Figure~\ref{fig:fig4}, right).
We also found no statistically significant interaction between catchment size and agriculture.

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figure4.pdf}
  \caption{Effect of agriculture within the catchment (left) and catchment size (right) on the number of RAC exceedances. Red line marks statistically significant changes. Dashed lines denote 95\% point-wise Confidence Intervals.
  }
  \label{fig:fig4}
\end{figure}

>>>>

%% -------------------------------------------------------------------------
\section{Methods}
\subsection{Data compilation}
We compiled pesticide monitoring data from sampling sites with catchment sizes $\mathrm{< 150km^2}$ for the years 2005 to 2015 from all 13 non-city federal states of Germany (see Supplemental Table~S1 for the abbreviations of federal state names). 
We homogenized and unified all data provided by the federal states into a database and implemented a robust data cleaning workflow (see Supplemental Figure~S1 for details) \citep{poisot_best_2015}.

We identified chemical samples taken during heavy rainfall events by spatio-temporal intersection of sampling events with gridded daily precipitation data available from the German Weather service (DWD).
%%% add resolution 
This data spatially interpolates daily precipitation values from local weather stations \citep{rauthe_central_2013}. 
We performed the intersection for the actual sampling date and the day before, to identify rain events during and up to 48 hours before sampling. 
%%%! One question with such data is from which precipitation value increases in concentrations are detectable. Also interesting

\subsection{Characterization of catchments}
We compiled a total of 3,049 sampling sites with pesticide measurements.
We delineated catchments upstream for each of the sampling sites using a digital elevation model (DEM) \citep{eea_digital_2013} and the multiple flow direction algorithm \citep{holmgren_multiple_1994} as implemented in GRASS GIS 7 \citep{neteler_grass_2012}.
Catchment delineation was manually checked for accuracy by comparison with state stream networks.
Given that the delineated catchments were accurate only for 30\% of the sites, we compiled complementary catchment size data from authorities that provided catchments for an additional (60\% of sites). 
Ten percent of sites lacked catchment size data and were omitted from the operations outlined below.

For each derived catchment (either from DEM or drainage basins) we calculated the relative cover (in \%) with agricultural area based on Official Topographical Cartographic Information System (ATKIS) of the land survey authorities \citep{adv_atkis_2016}. 
Additionally, we used agricultural cover data provided by authorities (19\% of sites). 
For 78\% of the sites both, the proportion of agricultural land use and catchment size were available.
% see do_overview.R for numbers

A clear definition of SWB in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
The WFD defines SWB with a catchment size between 10 and 100 km\textsuperscript{2}, without further categorisation of streams \textless 10km\textsuperscript{2}. 
\citet{lorenz_specifics_2016} defines SWB with catchment size \textless 10km\textsuperscript{2} as SWB.
Because of data scarcity of streams \textless 10~km\textsuperscript{2} (Figure~\ref{fig:fig3}) we define in this study all streams below 25~km\textsuperscript{2} as SWB. This catchment size corresponds to a stream width of approximately 2~meters (see Supplemental Figure~S2).

\subsection{Characterization of pesticide pollution}
We characterised pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorization as part of the ecological risk assessment.
No unacceptable ecological effect are expected if the environmental concentration remains below this concentration.
The German Federal Environmental Agency (UBA) provided RACs for the 105 compounds with highest detection rates (Supplemental Table~S2). 
We expressed RACs as Risk Quotient (RQ):

@@

where $No_i$ is the observed number of RAC exceedances at site $i$. 
We modeled $No_i$ as resulting from a negative binomial distribution ($NB$).
The proportion of agriculture within the catchment ($agri_i$) and the catchment size of the site ($size_i$) were used as predictors. 
$f_1$ and $f_2$ are smoothing functions using penalized cubic regression splines \citep{wood_generalized_2006}.
The degree of smoothness was estimated using restricted maximum likelihood (REML) during model fitting process \citep{wood_fast_2011}.
The number of samples per site ($n_i$) was used as an offset to account for differences in sampling efforts (sampling interval and analysed compound spectrum) at a site and is equivalent to modeling the rate of exceedances. 
We used point-wise 95\% Confidence Intervals (CI) of the first derivative of the fitted smooth to check if there are regions of statistically significant changes.
GAMs were fitted using the mgcv package \citep{wood_fast_2011}.

While agricultural land use and catchment size vary only between sites, this is not the case for precipitation which changes also with time.
Therefore, we modeled the effects of precipitation in a separate model.
RQ and concentrations show a skewed distribution with an excess of zeros (no pesticides detected and quantified). 
Therefore, we modeled these as two processes (one generating values below the limit of quantification (LOQ) and one generating values above LOQ) using a Zero-Adjusted Gamma (ZAGA) distribution \cite{rigby_generalized_2005,stasinopoulos_gamlss.dist:_2016}:

@@

$\nu_i$ denotes the probability of an observation i being above LOQ and $f_{Gamma}$ denotes the gamma function and is used for values equal to or greater LOQ, with $\mu$ being the mean and $\sigma$ the standard deviation of RQ.
We used the $log(x+0.05)$ transformed precipitation at sampling date ($log~prec_0$) and the day before ($log~prec_{-1}$), as well as quarters of the year ($Q1-Q4$) as linear predictors for $\mu$ and $\nu$. 
We used appropriate link functions for $\mu$ and $\nu$ and assumed $\sigma$ to be constant. 
Equation~\ref{eqn:eqn4} summarises the deterministic part of the model.

@@

We fitted this model separately to each compound with a RAC, measured in least 1000 samples and with more than 5\% of values above LOQ (n = 24 compounds, see Supplemental Table~S3 for a list of compounds). 
To summarise the coefficients across the modeled compounds we used a random effect meta-analysis \citep{harrison_getting_2011}.

@@

The compiled dataset used in analysis comprised 2,918,604 pesticide measurements of 42,236 samples in 3,049 sampling sites.  %see do_overview.R for numbers.
These samples were taken via grab sampling.  
Composite samples of different durations that were taken in 33 sites were not considered in analysis, given a potential methodical bias and the low sample size.
We found large differences in the number of sampling sites between federal states and their spatial distribution (Figure~\ref{fig:fig1} and Supplemental Table~S1).

@@

The distribution of sampling sites across catchment sizes indicated a disproportionally low number of sites in catchments below $10~km^2$, which deviated from the expected distribution \citep{nadeau_hydrological_2007}.
%%%! check expected distribution (exponential distribution?)
Most sampling sites had catchment sizes between 10 and 25 $km^2$ (Figure~\ref{fig:fig3}). 
The proportion of agriculture in the catchments tended to decrease with increasing catchment size.

@@

\subsection{Influence of agricultural land use and catchment size}
The number of RAC exceedances increased strongly and statistically significant up to 25\% agriculture within the catchment.
Above this threshold the exceedances levelled followed by an increase above 75\% (Figure~\ref{fig:fig4}, left).
Catchment size had no statistically significant effect on the number of RAC exceedances (Figure~\ref{fig:fig4}, right).
We also found no statistically significant interaction between catchment size and agriculture.

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figure4.pdf}
  \caption{Effect of percent agriculture within the catchment (left) and catchment size (right) on the number of RAC exceedances. Red line marks statistically significant changes. Dashed lines denote 95\% point-wise Confidence Intervals.
  }
  \label{fig:fig4}
\end{figure}
